---
output: html_document
---
Course Project Data Analysis & Statistical Inference by Philipp Paier
==========================================

# Introduction
In this document, I want to discuss, if age of a Titanic passenger is related to the fact, if the passenger survived or died during the accident. This is motivated by the usual saying, that children and women have to be saved first in case of an incident on a ship. While I pick this specific incident and will do inference for this case and for age only, everyone should be interested in this question in general, to see if this saying is really something, that is also practiced. Also it intuitively seems less likely, that older people are able to survive such an accident. I again want to note, that only the Titanic accident will be covered and therefore a generalization to all accidents involving ships with casualties, won't be covered in this document.

**Notes:**  
Code has been developed and tested with **R 3.1.2** on **Windows 7**.

# Data
## How is the Data collected?
The data that will be discussed has been collected from the passenger list of the Titanic. The original passenger list of the Titanic consisted of 2224 people, of whom 1502 died during the accident. Kaggle - an online website for machine learning - provides a sample of 891 people of that passenger list with various information, including passengers age and survival status in the end. This set of 891 people will be the basis for my inferential study.

**Important NOTE:**
As 891 samples in the kaggle set is clearly more than 10% of the complete passenger list, we can't assume that this set is independent. Especially family members are likely to stick together. Thus it would not be feasible to make a Hypothesis Test or Confidence Interval based on that sample. Therefore I further sample 200 out of the 891 Kaggle samples randomely. I justify this to still be a random sample of the original entire population (the complete Titanic passenger list), because I assume that the original Kaggle sample is also taken randomely. This sample of 200 passengers - which I will call the **titanic200** from now on - now consists of less of 10% of the entire population, which makes the independence assumption feasible.

Here is the R-Code, that shows how the titanic200 sample is generated:
```{r loaddata, echo=TRUE, cache = FALSE}
## read data & take only age and survival status as variables
KaggleSampleAllVars <- read.csv("titanic_sample.csv")
KaggleSample <- data.frame(Age=KaggleSampleAllVars$Age, Survived=as.factor(KaggleSampleAllVars$Survived))
```

```{r loaddata2, echo=FALSE, cache = FALSE}
KaggleSample <- KaggleSample[complete.cases(KaggleSample),]
set.seed(14)
```

```{r randomsample, echo=TRUE, cache = FALSE}
## random sample 200 passengers
idx <- sample(seq(1,nrow(KaggleSample)),200)
titanic200 <- KaggleSample[idx,]

survivors  <- titanic200[titanic200$Survived==1,]
casualties <- titanic200[titanic200$Survived==0,]
```

## Units of Observations
The units of observations will be passengers of the Titanic. 

## Variables of Interest
The variables of interest in this document will be:
Age - the age of a passenger (numerical, discrete)
Survived - 1 if a passenger survived and 0 otherwise (categorical with 2 levels)

## What type of Study?
It is an obervational study, as it is only observed what happened to the passengers, without interfering. Only an association between the variables and the outcome is studied. It is also a retrospective study, as past data is used.

## Scope of Inference (Generalizability & Causality)
The population of interest are the passengers of the Titanic. As this is only an obervational study, only correlation and not necessary causalization can be derived. The findings of this analysis could be generalized to the entire population of the Titanic, but no causalisation can be derived and there might be sources of bias: It might be, that Kaggle didn't choose the sample of 891 passengers out of the complete passenger list randomely, but for the sake of simplicity we assume they did. If this assumption would be untrue, the findings of this document wouldn't generalizable to the entire passenger list.

# Exploratory Analysis
```{r, echo = FALSE, results = 'hide'}
require('knitr')
options(scipen=5)
opts_chunk$set(echo=FALSE, fig.height=5,fig.width=5)
```

## Summary Statistics & Visualization
For exploratory analysis we first calculate two sample statistics. Namely the mean & median age for survivors as well as for casualties:
```{r means, echo=TRUE, cache = FALSE}
meanAgeSurvivors  <- mean(survivors$Age)
meanAgeCasualties <- mean(casualties$Age)

medianAgeSurvivors  <- median(survivors$Age)
medianAgeCasualties <- median(casualties$Age)
```

The result shows, that the mean age for survivors in our sample is **`r meanAgeSurvivors`** and the 
mean age for casualties is **`r meanAgeCasualties`**. The median age for survivors is **`r medianAgeSurvivors`** and the median age for casualties is **`r medianAgeCasualties`**.
This indicates, that age might really differ in those two groups in general and age and survival status are correlated. However, inference is necessary to back up that assumption.

A Box-Plot shows similar insight:
```{r boxplot, echo=TRUE, cache = FALSE}
par(mfrow = c(1, 1))
boxplot(Age~Survived,data=titanic200)
```

It can be seen that the age of casualties seem to be spread more and the median age is also higher for casualties than for survivors.

Histograms for each category show similar results:
```{r hist, echo=TRUE, cache = FALSE}
par(mfrow = c(1, 2))
hist(survivors$Age, main ="Survivors")
hist(casualties$Age, main ="Casualties")
```

Though there are overall more casualties than survivors in our titanic200 set, the number of survivors, that are younger than 10 years old is bigger than the number of casualties younger than 10 years, while for all other age categories the number is smaller.
 
## Discussion
As in our sample of 200 titanic passengers, the mean age and median age differ by a few years, we get the feeling, that this might no be by chance but can be backed up with statistical inference. Also the boxplot and the histograms indicate, that survivors were on average younger than casualties. This motivates the hypothesis test and confidence interval calculated in the next section.

# Inference
## Hypotheses
In this section we want to do a hypothesis test and calculate a confidence value, to see if difference in **mean age** has happened just by chance, or if it really indicates that survivors were on average younger than casualties. We will do this, by inference on the difference of the two means.

Consequently the Null-Hypothesis is, that the difference between the mean ages is 0, while the alternative hypothesis is that meanAgeSurvivors-meanAgeCasualties < 0.

## Conditions
**Independence:**
- as we assume, that Kaggle chose the passengers randomely and we also took ar ondom sample of the Kaggle test, we assume the random sampling condition is met
- as the 200 samples in our set is less than 10% of the more than 2000 passengers of the titanic, the 10% rule holds true
- as we assume, that the smaples have been taken randomely, we don't expect dependence between the two groups

**Sample Size/Skew:**
As in our data set there are 83 survivors and 117 casualties, both our groups are larger than 30. Also the histograms shown earlier, don't look too skewed, but with more than 80 samples for each group we should be on the save side anyway.

All conditions hold true, thus allowing to do inference comparing the two means.

## Used methods
We will calculate a 95% confidence interval for the difference in the two means as well as doing a one-sided hypothesis test with alpha being 0.025. As for a one sided hypothesis test a threshold of 2.5 is equivalent two a confidence interval of 1-2*0.025 = 0.95, we can compare the two results in the end. 

# Confidence Interval
To calcualte the 95% confidence interval, we first have to calculate the difference of mean age and standard error:
```{r ci1, echo=TRUE, cache = FALSE}
## calculate difference in mean age
dma <- meanAgeSurvivors - meanAgeCasualties

  ## calculate the standard error
sds <- sd(survivors$Age) 
sdc <- sd(casualties$Age)
se  <- sqrt(sds^2/nrow(survivors)+sdc^2/nrow(casualties))
```
The difference in mean age is **`r dma`** and the standard error is **`r se`**
This allows us to compute the 95% confidence interval:
```{r ci2, echo=TRUE, cache = FALSE}
## calculate difference in mean age
cilow  <- dma - 1.96*se 
cihigh <- dma + 1.96*se 
```
The confidence interval is (**`r cilow`**,**`r cihigh`**) and thus not including the null value 0.

## Interpretation of CI
The interpretation of the Confidence interval is as follows: We are 95% confident that the age of survivors of the titanic accident is on average **`r -cihigh`** to **`r -cilow`** lower than the age of casualties. 

As the Confidence Interval doesn't include the 0, this indicates, that survivors were in fact younger than casualties.

# Hypothesis Test
We compare the Null-Hypothesis, that there is no difference in mean age with the alternative one sided hypothesis, that the mean age of casualties is higher than that of survivors.

We compute the Z-Score and the p-value 
```{r ht, echo=TRUE, cache = FALSE}
## calculate the z score
z <- (dma-0)/se 
p <- pnorm(z)
```

The Z-Score is **`r z`** and the p-value is **`r p`**. Thus with a threshold of 0.025, we decide to reject the Null-hypothesis.

## Interpretation of p-Value & HT
The p-value **`r p`** is interpreted as follows. The likelihood of observing a difference of mean ages of **`r dma`**, while in fact the true difference is 0, is only **`r 100*p`**%.

In a one-sided hypothesis test with a threshold of 0.025 we therefore decide to reject the Null-Hypothesis, meaning that our sample delivered convincing evidence, that in the entire population survivors were on average in fact younger, than casualties.

# Do results agree?
As the we reject the Null-Hypothesis, because of the small p-value, and the Confidence Interval also doesn't include the Null-Value, the results of the two calculations agree with eachother. 

# Conclusion
By doing a Confidence Interval and a Hypothesis test we came to the conclusion, that "it's very likely" (we are 95% confident) that survivors of the titanic accident were on average younger than the casualties. This means, that there is a relationship between surviving and age, as the saying "Children and Women should be saved first" suggests. However, as this is an observational study, we can not make an causal statements and are thus not allowed to say, that being younger made it more likely to survive the titanic accident.

Another very important point is, that we assumed that Kaggle chose the people in there data set randomely. This might not be true. As a consequence we wouldn't be able to generalize our findings to the entire passenger list of the titanic, as our random sample of 200 people would be biased towards a non random sample.

Future work might involve including more variables by doing either inference for specific pairs of variables, or by grouping passengers into various groups, like e.g. female children, male children, male adults, female adults and so on. This would would mean that different approaches for inference are necessary, but might give insight, how other variables interact with eachother too.


# References
The original Titanic dataset can be found at Kaggle:
https://www.kaggle.com/c/titanic-gettingStarted/
You have to register to download the data. And only the train dataset can be used for this study, as for the test set, no surviving information is given. 

# Appendix
First 30 samples of the Titanic200 Sample:
```{r titanic200, echo=TRUE, cache = FALSE}
titanic200[seq(1,30),]
```
